---
title: "Summary Statistics & R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 1

### Instructions Week 1

You can go through the [swirl](http://swirlstats.com/) tutorial, which teaches you R programming and data science interactively, at your own pace and in the R console. Once you have R installed, you can install `swirl` and run it the following way: 

```{r, eval=FALSE}
install.packages("swirl")
library(swirl)
swirl()
```

You can download packages with the `install.packages` function and load them with the `library` function. You need to install the packages only once. Every session you need to load the packages into your library.

```{r, eval=FALSE}
install.packages("rafalib")
library(rafalib)
```

You can see your working directory by typing:

```{r, eval=FALSE}
getwd()
```

If you have this file in your working directory, then you can simply read it in like this:

```{r,eval=FALSE}
dat <- read.csv("femaleMiceWeights.csv")
```


To install packages from GitHub you will need to install the `devtools` package:

```{r,eval=FALSE}
install.packages("devtools")
```


### Exercizes Week 1


#### Some basic functions in R

Getting started with vectors and such. 

```{r,eval=FALSE}
x <- c(2.23, 3.45, 1.87, 2.11, 7.33, 18.34, 19.23)
mean(x)

sum <- 0
for(i in 1:25)
  sum <- sum + i^2
sum

# The cars dataset is available in base R.
class(cars)
nrow(cars)
mean(cars[,2])
which(cars[,2]==85)
```

Read in the file `femaleMiceWeights.csv` and report the exact name of the column containing the weights.

```{r,eval=FALSE}
dat <- read.csv("femaleMiceWeights.csv")
head(dat)
names(dat)[2]
```

What is the entry in the 12th row and second column?

```{r,eval=FALSE}
dat[12,2]
```

Use $ to extract the weight column and report the weight of the mouse in the 11th row.

```{r,eval=FALSE}
weights <- dat$Bodyweight
weights[11]
```

How many mice are included in our dataset?

```{r,eval=FALSE}
weights <- dat$Bodyweight
length(weights)
```

What is the average weight of mice on the high fat diet?

```{r,eval=FALSE}
View(dat) 
weights <- dat$Bodyweight
mean(weights[13:24])
```

Now take a random sample of size 1 from the numbers 13 to 24 and report back the weight of the mouse represented by that row. 

```{r,eval=FALSE}
set.seed(1)
i <- sample(13:24, 1)
dat$Bodyweight[i]
```


#### Using dplyr package

For the next exercises, we will use a new dataset related to mammalian sleep. 

```{r, eval=FALSE}
dat <- read.csv("msleep_ggplot2.csv")
class(dat)
```

We will use the `dplyr` functions, `select()` and `filter()`. Now use the `filter()` function to select only the primates. How many animals in the table are primates?

```{r, eval=FALSE}
library(dplyr)
head(dat)
dat2 <- filter(dat, order=="Primates")
nrow(dat2)
class(dat2)
```

Now we want to calculate the average amount of sleep for primates (the average of the numbers computed above). One challenge is that the `mean()` function requires a vector so, if we simply apply it to the output above, we get an error. Use `unlist()` and to compute the desired average. What is the average amount of sleep for primates?

```{r, eval=FALSE}
y <- filter(dat, order=="Primates") %>% select(sleep_total) %>% unlist
mean(y)

filter(dat, order=="Primates") %>% summarize( mean( sleep_total) )  # same but shorter
```


#### Plots in R

Load new data into R
```{r, eval=FALSE}
load("skew.RData")
dim(dat)
```

You can use the following line of code to set up a grid for 3x3=9 plots. (`mfrow` means we want a multifigure grid filled in row-by-row.Another choice is `mfcol`.)

```{r, eval=FALSE}
par(mfrow = c(3,3))
```

Which column has positive skew (a long tail to the right)?

```{r, eval=FALSE}
for (i in 1:9) {
    qqnorm(dat[,i])
    qqline(dat[,i])
}
hist(dat[,4])

```

Which spray seems the most effective (has the lowest median count)? --> C

```{r, eval=FALSE}
head(InsectSprays)
boxplot(split(InsectSprays$count, InsectSprays$spray))

boxplot(InsectSprays$count ~ InsectSprays$spray)  # same but shorter
```


Use boxplots and histograms to compare the finishing times of males and females. Which of the following best describes the difference?

```{r, eval=FALSE}
library(dplyr)
data(nym.2002, package="UsingR")
par(mfrow = c(1,3))
males <- filter(nym.2002, gender=="Male") %>% select(time) %>% unlist
females <- filter(nym.2002, gender=="Female") %>% select(time) %>% unlist
boxplot(females, males)
hist(females,xlim=c(range( nym.2002$time)))
hist(males,xlim=c(range( nym.2002$time)))
```

